### 2019-10-09 submission/manuscript.tex
Incorporate MB / KS comments

More than 10 years ago, we published the manuscript describing the
mothur software package in \emph{Applied and Environmental
Microbiology}. Our goal was to create a comprehensive package that
allowed users to analyze amplicon sequence data using the most robust
methods available. Looking back, it is clear that mothur continues to
provide this service to the microbial ecology community as it has helped
to lead the community through the ongoing sequencing revolution. Beyond
its success and impact on the field, mothur's development exposed a
series of observations that are generally translatable across science.
Although it is easy to criticize choices that were made 10 years ago
through a modern lens, all science is done in the context of prevailing
ideas and available technologies. If we were to wait for all of the
possible limitations to be solved before proceeding, science would
stall. Even preceding the development of mothur, it was necessary to
address the most important problems and work backwards to other problems
that limited access to robust sequence analysis tools. At the same time,
we strive to expand mothur's capabilities in a data-driven manner to
incorporate new ideas and accommodate changes in data and desires of the
research community. As we develop mothur, we first consider the needs of
my research group. In hindsight, we appreciate that our needs are
representative of the broader community's. It has been edifying to see
the benefit that a simple set of tools can bring to so many other
researchers.

@@

Looking back on scientific journeys can be instructive to others who are
overwhelmed at the prospect of looking forward at their careers (1--5).
By no means is my personal journey over, but since 2002 I have been on a
journey that I did not realize I was on. Now that the paper introducing
the mothur software package is ten years old and has become the most
cited paper published by \emph{Applied and Environmental Microbiology}
(7), it is worth stepping back and using the continued development of
mothur as a story that has parallels to many other research stories.

@@

It is important to remember that we knew there were many problems with
16S rRNA gene sequencing. We knew there were biases from extractions and
amplification (18--23). We knew there were chimeras (24--27). We knew
that bacteria varied in their \emph{rrn} copy number. Generating a
distance matrix was a prerequisite to using my tools. This wasn't
trivial, but by cobbling together other tools it was possible. We would
assemble, trim and and correct Sanger sequence reads using Chromas or
STADEN (28), align the sequences using ClustalW (29) or ARB (30), check
for chimeras using partial treeing or Bellerephon (27), and calculate a
pairwise distance matrix using DNADIST from the PHYLIP package (31). At
the time, we knew that we only had a loose concept of a species based on
these distances (32). We hoped that an OTU defined as a group of
sequences more than 97\% similar to each other would be a biologically
meaningful unit regardless of whether it fit our notion of a bacterial
species. At the time, I felt that the biggest problems that I could
solve were how to cluster the sequences into OTUs and how to use those
clusterings to test our hypotheses. The only tool available at the time
that automated the clustering step was FastGroup, which implemented a
greedy version of the single linkage algorithm (33). The high cost of
sequencing was also an impediment to experimentation and analysis in
microbial ecology. It was rare for a study design to have experimental
replicates so that one could perform a statistical test to compare
treatment groups. For example, in our testing we frequently used a
dataset comparing Scottish soils from from Alison McCaig and colleagues
(34). This dataset consisted of two experimental groups, each replicated
three times with 45 sequences per replicate. Although great focus has
been placed on the depth of sampling afforded by 454 and Illumina
sequencing, the true benefit of the modern sequencing platforms is the
ability to affordably sequence a large number of technical and
biological replicates. In spite of the many technical challenges, we had
excuses and heuristics to solve problems that served our needs. It is
telling that a recent review of ``best practices'' in generating and
analyzing 16S rRNA gene sequences shows that we still have not solved
many of these issues and that we have even identified additional
problems (35).

As we developed these tools, I found a unique niche in microbiology. My
undergraduate and graduate training as a biological engineer prepared me
to think about research questions from a systems perspective, to think
quantitatively, and to understand the value of using computer programs
to help solve problems. As an undergraduate student, I learned the
Pascal programming language and promptly forgot much of it as a graduate
student and learned MatLab in its place. As a postdoc, I learned the
Perl programming language to better understand how LIBSHUFF, a tool for
comparing the structure of two communities, worked since it was written
in Perl (36). After writing my own version of LIBSHUFF,
\(\int\)-LIBSHUFF and seeing the speed of the version written in C++ by
my collaborator, Bret Larget, I converted my Perl version of DOTUR into
C++. At the time, the conversion from Perl to C++ seemed like an
academic exercise to learn a new language. My Perl version only took a
minute or so to process the final collection of 1,000 sequences and the
C++ version took seconds. Was that really such a big difference? In
hindsight, as we now process datasets with millions of sequences, the
decision to learn to C++ was critical. The ability to pick up computer
languages to solve problems was enabled by my prior training. It was
also a skill that was virtually unheard of in microbiology. Today,
researchers without the ability to program are at a significant
disadvantage (37).

@@

Milling about at a poster session at an ASM General Meeting in New
Orleans, I ran into Mitch who asked what my plans were for my new lab. I
told him that I wanted to make a tool like ARB, a powerful database tool
and phylogenetics package (30), but for microbial ecology analysis. His
retort was, ``You and what army?'' Up to that point, I had written every
line of code and been answering many emails from people asking for help.
He was right, I would need an army. It would be difficult, but I needed
to learn to let go and share the development process with someone else.
My ``army'' ended up being Sarah Westcott who has worked on the mothur
project from its inception. Today, mothur is over 200,000 lines of code
and Sarah has touched or written nearly every line of it. Beyond writing
and testing mothur's code base, she has become a conduit for many who
are trying to learn the tools of microbial ecology. She patiently
answers questions via email and on the package's discussion forum
(\url{https://forum.mothur.org}). The community and I are lucky that
Sarah has stayed with the project for more than a decade. To be honest,
such dependency on a single person makes the project brittle. In
hindsight, it would have been better to have developed mothur with more
of an ``army'' or team so that there is overlap in people's
understanding of how mothur works. Although a distributed team approach
might work in a software engineering firm, it is not practical in most
academic environments where there is limited funding. There are
certainly projects that make this work, but they are rare.

@@

From the beginning there have been online tools available at the
Ribosomal Database Project (RDP) (39), greengenes (40), and SILVA (41).
These allowed users a straightforward method of comparing their data to
those collected in a database. There are two primary downsides to these
tools. First, researchers running the online tool must pay the
computational expenses leading to slow process times when hardware
becomes outdated and when numerous users simultaneously attempt their
analysis. Eventually this limitation would result in the termination of
the greengenes website. Second, these platforms provide a
one-size-fits-all analysis. These tools only allow a user to analyze 16S
and in some cases 18S rRNA gene sequences. If a user sequences a
different gene, then the tool will not serve them. These observations
resulted in two design goals we have had with mothur: bringing the
analysis to a user's computer and separating a tool from a specific
database. For example, we commonly use a sequence alignment method that
was originally developed for greengenes (42), but use a SILVA-based
reference alignment because its superior quality (43, 44). In addition,
we offer the \text{na\"ive} Bayesian classifier developed by the RDP
(45) and allow users to train it to any database they want, including
customized databases. In both examples, users can align or classify
non-rRNA gene sequence data. As the bioinformatics tools have matured,
both the RDP and SILVA offer integrated pipelines for analyzing large
datasets, albeit in one-size-fits-all black box implementations.

With the growth in popularity of 16S rRNA gene sequencing there has
naturally been an expansion in the number of people developing tools to
analyze these data. Months after the paper describing mothur was
published, the paper describing QIIME was published (46). Over the past
10 years, many have attempted to create analogies comparing the two
programs: Pepsi vs Coke, Apple vs Windows, etc. It is never clear which
software is which brand and whether the comparisons are meant as a
complement or an insult. Regardless, both programs are very popular.
From my perspective, most of the differences are cosmetic. QIIME is
effectively a bundle of scripts to run other developers' software. For
example, with QIIME (through version 1.9.1), it was even possible to run
mothur through QIIME. One can also run the \text{na\"ive} Bayesian
classifier through QIIME using the original code developed by the RDP.
This caused great frustration for many users because there were numerous
software dependencies that had to be installed. Although the QIIME
developers would go on to create virtual machines and use packaging
tools to simplify installation, these fixes required sophistication by
users who we knew struggled with the basics of navigating a command
line. In contrast, when a user runs mothur, they are running mothur. The
\text{na\"ive} Bayesian classifier code that is in mothur is a rewritten
version of the original code. When we rewrite someone's software we do
it with an eye to improving performance, access, and utility for non-16S
rRNA gene sequence data. For example, while 454 data was popular,
PyroNoise was an effective tool for denoising flowgram data (47).
Running the original code required a large Linux computer cluster and
knowledge of bash and Perl scripting. When we rewrote the code for
mothur, we made it accessible to people using any operating system with
a simple command interface (i.e.~trim.flows and shhh.flows). Our
approach requires significant developer effort, but saves considerable
user effort. As this benefit is multiplied across thousands of projects,
the savings to users has been considerable.

Beyond the large packages like mothur and QIIME, there has been
significant growth in stand alone software tools for sequence curation
(e.g.~PyroNoise (47), PANDAseq (48), DADA2 (49)), chimera checking
(e.g.~UCHIME (50), ChimeraSlayer (51), Perseus (52)), and clustering
(e.g.~USEARCH (53), VSEARCH (54), Swarm (55)). Where possible and when
warrented, we have implemented many of these algorithms directly into
mothur. We have also used this diversity of methods to perform
head-to-head comparisons. Most notable is the area of clustering
algorithms where there have been a large number of algorithms developed
without an obvious method to objectively compare them (56--59). We
applied an objective metric, the Matthew's Correlation Coefficient
(MCC), to evaluate numerous algorithms for clustering sequences into
OTUs. By performing this type of analysis, we were able to objectively
compare the algorithms, make recommendations to the field, and develop
new algorithms that outperformed the existing algorithms. Beyond
evaluating clustering algorithms, we have also evaluated methods of
denoising sequence data (60--62), assessed reference alignments (43,
44), considered the importance of incorporating secondary structure
information in alignments (63), quantified the variation along the 16S
rRNA gene (44), and compared the statistical hypotheses tested by
commonly used tools (64). We have embraced the competition and diversity
of all methods being used to analyze amplicon data as an opportunity to
identify the strengths and weaknesses of the methods in an attempt to
make recommendations to other researchers.

@@

leverage existing tools for other steps. Yet, when we learned that the
code for NAST, the algorithm behind the greengenes's aligner (42), was
not open source or publicly available. Similarly, although the SINA
aligner was available through ARB and the SILVA website performed well
it was closed source. Because the ARB implementation did not scale to
large datasets, researchers were left to pay a processing fee to the
SILVA website to align sequences. Thus, we realized that such an
important tool needed to be opened to the community (43). More recently,
the rejection of closed source, commercial tools such as USEARCH can be
seen by the broader adoption of its open source, free competitor,
VSEARCH, within the microbial ecology community (53, 54). Related to
insuring that mothur's code is open source, our second core principle is
that we maintain transparency to our users. Perhaps a user does not need
to interrogate every line of code, but they need to understand what is
happening. Many programs including online workflows encapsulate large
elements of a pipeline in a single command. In contrast, mothur forces
the user to specify each step of the pipeline. Although the former
approach makes an analysis easier for a beginner, it potentially stifles
users that need greater control or understanding of the assumptions at
each step. This control over the pipeline has made it easier for
researchers to customize databases or adapt the pipeline to analyze
non-16S rRNA gene sequence data. Third, as I mentioned above, there has
been a plethora of methods proposed for generating amplicon sequence
data, and curating, aligning, checking for chimeras, classifying, and
clustering the data. I am proud of the data-driven approach we have
taken to comparing these methods. A description of a new method is of
limited value if it is not benchmarked against other methods or control
datasets. Through this core principle and mothur's large reach into the
community, we have helped to develop standards in the analysis of 16S
rRNA gene sequence data. Fourth, a focus on enabling reproducibility has
always been central to the functionality of mothur. From the beginning,
mothur's logfiles have represented a transcript of the user's command
and outputs. When researchers were reluctant to submit sequence data to

@@

create a mothur command (make.sra) to create the package to submit

@@

generate. We ``eat our own dogfood''. This has proven critical as it
again represents transparency and hopefully provides confidence to
mothur's users that we are not making recommendations that we do not
follow ourselves.

\textbf{\emph{Challenges of making open source count.}} Anyone can post
code to GitHub with a permissive license and claim to be an open source
software developer. Far more challenging is engaging the target
community to make contributions to that code. Frankly, we have struggled
to expand the number of people that make contributions to the mothur
code base. One challenge we face is that if we looked to third parties
to contribute code to mothur, they would need to know C++. Given the
paucity of microbiologists with skills programing in a compiled language
like C++, expecting that community to provide contributors that can
write code in a syntax that prizes execution efficiency over developer
efficiency was not likely. In contrast, the QIIME development team could
be more distributed because their code base was primarily written in
Python, which prizes developer efficiency over execution efficiency.
QIIME is a series of wrappers that allow users to execute other
developers' code making the use of a scripting language like Python
attractive. Their choices resulted in many tradeoffs that have impacted
ease of installation, usability, execution speed, and flexibility. If we
were offered funding to rewrite mothur, we would likely rewrite it as an
R package that leaned heavily on the R language's C++ interface
packages. Of course, such choices are always best in hindsight. Yet,
when we started developing mothur, the ability to interface between
scripting languages like R and Python and C++ code was not as well
developed as it is today. For example, the modern version of the Rcpp
package was first released in 2009 and its popularity was not immediate
(66). Again, the development of mothur has been a product of the
environment that it was created in. Although these decisions have
largely had positive outcomes, there have been tradeoffs that caused us
to sacrifice other goals.

Beyond contributing to the mothur code base, we sought out other ways to
include the community as developers. The paper describing mothur
included 15 co-authors, all but three (Schloss, Westcott, and Ryabin)
responded to a call to provide a wiki page that described how they used
an early version of mothur to analyze a data set. Our vision was that
authors might use the mothur wiki to document reproducible workflows for
papers using mothur but to also provide instructional materials for
other seeking to adapt mothur for their uses
(\url{https://www.mothur.org/wiki}). Unfortunately, once the incentive
of co-authorship was removed, researchers stopped contributing their
workflows to the wiki. Again, this vision and the lack of the
community's adoption of wikis as a mechanism for reporting workflows was
a product of the environment. Although wikis were popular in the late
2000's, they lacked the ability to directly execute the commands that
researchers reported. Such technology would not be possible until the
creation of IPython notebooks (2011) and R markdown (2012). Another
problem with the wiki approach was that potential contributors did not
see the wiki as a community resource. I frequently received emails from
scientists telling me that there was a typo on a specific page when the
intention was that they could correct the typos without my input. We
have been more successful in soliciting input and contributions from the
user community through the mothur discussion forum and GitHub-based
issue tracker. As mothur has matured, we have been dependent on the user
community to use these resources to tell us what features they would
like to see included in mothur and where the documentation is confusing
or incomplete (\url{https://forum.mothur.org}). Often we can count on
people not directly affiliated with mothur to provide instruction and
their own experience to other users on the forum. We are constantly
trying to recruit our ``army'' and are happy to take any contributions
we can. Whether the contributions are to the code base, discussion
forum, or suggestions for new tools, these contributions have been
invaluable to the growth and popularity of mothur.

\textbf{\emph{Failed experiments.}} If we never failed, we would not be
trying hard enough. Over the past decade we have tried a number of
experiments to improve the usability and utility of mothur. One of our
first experiments was to use mothur to generate standard vector graphic
(SVG)-formatted files of heatmaps and Venn diagrams depicting the
overlap between microbial communities. Such visuals were helpful for
exploring or data; however, I quickly realized that I would never put a
mothur-generated figure into a manuscript I wrote. Such visuals require
far too much customization to be publication-quality. Although QIIME has
incorporated visualization tools through the Emperor package (67), the
challenge of users taking default values has downsides as ordinations
with black background or publishing 3-D ordinations in a 2-D medium
litter the literature. Instead, we have encouraged users to use R
packages to visualize mothur-generated results using the minimalR
instructional materials that I have developed
(\url{http://www.riffomonas.org/minimalR/}). A second experiment was the
creation of a graphical user interface (GUI) for running mothur. Forcing
users to interact with mothur through the command line has been a
significant hurdle for many. Unfortunately, the development effort
required to create and maintain a GUI is significant and there is
limited funding for such efforts. The newest version of QIIME (starting
with version 2.0.0) has emphasized interaction with the tools through a
GUI (68) and the related QIITA project offers a web-based GUI (69). It
remains to be seen how this experiment will go. Another downside of
using a GUI is that there is a risk that reproducibility will suffer if
users do not have a mechanism to document their mouse clicks.
Documentation of commands and parameter values is explicit in mothur as
users can provide the software a file with a list of commands and all
commands and output are recorded in a logfile. Given the heightened
focus on reproducibility in recent years (70), we have extended
significant effort in developing instructional materials teaching users
how to organize, document, and execute reproducible pipelines that allow
a user to go from raw sequence data to a compiled manuscript with
figures through the Riffomonas project (65). A final example of a failed
experiment was a collaboration with programmers through Google Summer of
Code to develop commands in mothur that ran the random forest and SVM
machine learning algorithms. Similar to the challenges of developing
attractive visuals, fitting the algorithms' hyperparameters, testing,
and deploying the resulting models require a significant amount of
customization. Furthermore, machine learning is an active area of
research where methods are still being developed and improved.
Thankfully, there are numerous R and Python packages that do a better
job of developing these models (71, 72). Again, we have put our efforts
into developing instructional materials that mothur users can use to fit
such models to their data. In each of our ``failed'' experiments, the
real problems were straying from what mothur does well and failing to
grasp what we really wanted the innovation to do.

@@

The development of mothur would not be possible without the
contributions of its many supporters, developers, and users. Although
not a complete list, I would be remiss if I did not express my gratitude
to Sarah Westcott and the other members of my research group, Jo
Handelsman, Mitch Sogin, Susan Huse, Vincent Young, Lita Proctor, Kendra
Mass, and Marcy Balunas for their unique contributions to the continued
development of mothur.

>>>>

More than 10 years ago, we published the manuscript describing the
mothur software package in \emph{Applied and Environmental
Microbiology}. Our goal was to create a comprehensive package that
allowed users to analyze amplicon sequence data using the most robust
methods available. mothur has helped lead the community through the
ongoing sequencing revolution and continues to provide this service to
the microbial ecology community. Beyond its success and impact on the
field, mothur's development exposed a series of observations that are
generally translatable across science. Perhaps the observation that
stands out the most is that all science is done in the context of
prevailing ideas and available technologies. Although it is easy to
criticize choices that were made 10 years ago through a modern lens, if
we were to wait for all of the possible limitations to be solved before
proceeding, science would stall. Even preceding the development of
mothur, it was necessary to address the most important problems and work
backwards to other problems that limited access to robust sequence
analysis tools. At the same time, we strive to expand mothur's
capabilities in a data-driven manner to incorporate new ideas and
accommodate changes in data and desires of the research community. It
has been edifying to see the benefit that a simple set of tools can
bring to so many other researchers.

@@

Looking back on scientific journeys can be instructive to others who are
overwhelmed at the prospect of looking forward at their careers (1--5).
By no means is my scientific journey over, but since 2002 I have been on
a journey that I did not realize I was on. Now that the paper
introducing the mothur software package is ten years old and has become
the most cited paper published by \emph{Applied and Environmental
Microbiology} (7), it is worth stepping back and using the continued
development of mothur as a story that has parallels to many other
research stories.

@@

It is important to remember that we knew there were many problems with
16S rRNA gene sequencing. We knew there were biases from extractions and
amplification (18--23). We knew there were chimeras (24--27). We knew
that bacteria varied in their \emph{rrn} copy number. Generating a
distance matrix was a prerequisite to using my tools. This wasn't
trivial, but by cobbling together other tools it was possible. We would
assemble, trim and correct Sanger sequence reads using Chromas or STADEN
(28), align the sequences using ClustalW (29) or ARB (30), check for
chimeras using partial treeing or Bellerephon (27), and calculate a
pairwise distance matrix using DNADIST from the PHYLIP package (31). At
the time, we knew that we only had a loose concept of a species based on
these distances (32). We hoped that an OTU defined as a group of
sequences more than 97\% similar to each other would be a biologically
meaningful unit regardless of whether it fit our notion of a bacterial
species. At the time, I felt that the biggest problems that I could
solve were how to cluster the sequences into OTUs and how to use those
clusterings to test our hypotheses. The only tool available at the time
that automated the clustering step was FastGroup, which implemented an
approximation of the single linkage algorithm (33). The high cost of
sequencing was also an impediment to experimentation and analysis in
microbial ecology. It was rare for a study design to have experimental
replicates so that one could perform a statistical test to compare
treatment groups. For example, in our testing we frequently used a
dataset comparing Scottish soils from Alison McCaig and colleagues (34).
This dataset consisted of two experimental groups, each replicated three
times with 45 sequences per replicate. Although great focus has been
placed on the depth of sampling afforded by 454 and Illumina sequencing,
the true benefit of the modern sequencing platforms is the ability to
affordably sequence a large number of technical and biological
replicates. In spite of the many technical challenges, we had excuses
and heuristics to solve problems that served our needs. It is telling
that a recent review of ``best practices'' in generating and analyzing
16S rRNA gene sequences shows that we still have not solved many of
these issues and that we have even identified additional problems (35).

As we developed these tools, I found a unique niche in microbiology. My
undergraduate and graduate training as a biological engineer prepared me
to think about research questions from a systems perspective, to think
quantitatively, and to understand the value of using computer programs
to help solve problems. As an undergraduate student, I learned the
Pascal programming language and promptly forgot much of it. Although it
was a good language for teaching programming concepts, it did not catch
on outside of the classroom. Later, I learned MATLAB. Because it was an
expensive commercial programming environment and never caught on with
biologists, I also forgot much of it. Even if I forgot the programming
syntax of these languages, what these programming experiences taught me
was the logic and structure of programming. As a postdoc, I would use
these experiences to learn the Perl programming language to better
understand how LIBSHUFF, a tool for comparing the structure of two
communities, worked since it was written in Perl (36). After writing my
own version of LIBSHUFF, \(\int\)-LIBSHUFF, and seeing the speed of the
version written in C++ by my collaborator, Bret Larget, I converted my
Perl version of DOTUR into C++. At the time, the conversion from Perl to
C++ seemed like an academic exercise to learn a new language. My Perl
version only took a minute or so to process the final collection of
1,000 sequences and the C++ version took seconds. Was that really such a
big difference? In hindsight, as we now process datasets with millions
of sequences, the decision to learn C++ was critical. The ability to
pick up computer languages to solve problems, enabled by my prior
training in engineering, was a skill that was virtually unheard of in
microbiology. Today, researchers without the ability to program are at a
significant disadvantage (37).

@@

Milling about at a poster session at the 2007 ASM General Meeting in
Toronto, I ran into Mitch who asked what my plans were for my new lab. I
told him that I wanted to make a tool like ARB, a powerful database tool
and phylogenetics package (30), but for microbial ecology analysis. His
retort was, ``You and what army?'' Up to that point, I had written every
line of code and been answering many emails from people asking for help.
He was right, I would need an army. It would be difficult, but I needed
to learn to let go and share the development process with someone else.
My ``army'' ended up being Sarah Westcott who has worked on the mothur
project from its inception. Today, mothur is over 200,000 lines of code
and Sarah has touched or written nearly every line of it. Beyond writing
and testing mothur's code base, she has become a conduit for many who
are trying to learn the tools of microbial ecology. She patiently
answers questions via email and on the package's discussion forum
(\url{https://forum.mothur.org}). The community and I are lucky that
Sarah has stayed with the project for more than a decade. To be honest,
such dependency on a single person makes the project brittle. In
hindsight, it would have been better to have developed mothur with more
of an ``army'' or team so that there is overlap in people's
understanding of how mothur works. Although a distributed team approach
might work in a software engineering firm, it is not practical in most
academic environments where there is limited funding. There are
certainly projects that make this work, but they are rare.

@@

From the beginning there have been online tools available at the
Ribosomal Database Project (RDP) (39), greengenes (40), and SILVA (41).
These allowed users a straightforward method of comparing their data to
those collected in a database. There are two primary downsides to these
tools. First, researchers running the online tool must pay the
computational expenses. When their hardware becomes outdated because it
is expensive to replace or maintain, processing times slow down.
Eventually this limitation would result in the termination of the
greengenes website. Second, these platforms provide a one-size-fits-all
analysis. These tools only allow a user to analyze 16S and in some cases
18S rRNA gene sequences. If a user sequences a different gene, then the
tool will not serve them. These observations resulted in two design
goals we have had with mothur: bringing the analysis to a user's
computer and separating a tool from a specific database. For example, we
commonly use a sequence alignment method that was originally developed
for greengenes (42), but use a SILVA-based reference alignment because
its superior quality (43, 44). In addition, we offer the \text{na\"ive}
Bayesian classifier developed by the RDP (45) and allow users to train
it to any database they want, including customized databases. In both
examples, users can align or classify non-rRNA gene sequence data. As
the bioinformatics tools have matured, both the RDP and SILVA offer
integrated pipelines for analyzing large datasets, albeit in
one-size-fits-all black box implementations.

With the growth in popularity of 16S rRNA gene sequencing there has
naturally been an expansion in the number of people developing tools to
analyze these data. Months after the paper describing mothur was
published, the paper describing QIIME was published (46). Over the past
10 years, many have attempted to create analogies comparing the two
programs: Pepsi vs Coke, Apple vs Windows, etc. It is never clear which
software is which brand and whether the comparisons are meant as a
complement or an insult. Regardless, both programs are very popular.
From my perspective, most of the differences are cosmetic
(\url{http://blog.mothur.org/2016/01/12/mothur-and-qiime/}). To me the
most meaningful difference between mothur and QIIME is the choice of
algorithms used to cluster sequences into OTUs. QIIME's advocacy for
open and closed-reference clustering and USEARCH or VSEARCH-based
\emph{de novo} clustering results in lower quality OTU assignments
relative to the \emph{de novo} clustering algorithms available within
mothur (47, 48). QIIME is set of wrapper scripts that help users to
transition data between independent packages. For example, with QIIME
(through version 1.9.1), it was even possible to run mothur through
QIIME. One can also run the \text{na\"ive} Bayesian classifier through
QIIME using the original code developed by the RDP. This caused great
frustration for many users because there were numerous software
dependencies that had to be installed. Although the QIIME developers
would go on to create virtual machines and use packaging tools to
simplify installation, these fixes required sophistication by users who
we knew struggled with the basics of navigating a command line. In
contrast, when a user runs mothur, they are running mothur. The
\text{na\"ive} Bayesian classifier code that is in mothur is a rewritten
version of the original code. When we rewrite someone's software we do
it with an eye to improving performance, access, and utility for non-16S
rRNA gene sequence data. For example, while 454 data was popular,
PyroNoise was an effective tool for denoising flowgram data (49).
Running the original code required a large Linux computer cluster and
knowledge of bash and Perl scripting. When we rewrote the code for
mothur, we made it accessible to people using any operating system with
a simple command interface (i.e.~trim.flows and shhh.flows). Our
approach requires significant developer effort, but saves considerable
user effort. As this benefit is multiplied across thousands of projects,
the savings to users has been considerable.

Beyond the large packages like mothur and QIIME, there has been
significant growth in standalone software tools for sequence curation
(e.g.~PyroNoise (49), PANDAseq (50), DADA2 (51)), chimera checking
(e.g.~UCHIME (52), ChimeraSlayer (53), Perseus (54)), and clustering
(e.g.~USEARCH (55), VSEARCH (56), Swarm (57)). Where possible and when
warranted, we have implemented many of these algorithms directly into
mothur. We have also used this diversity of methods to perform
head-to-head comparisons. Most notable is the area of clustering
algorithms where there have been a large number of algorithms developed
without an obvious method to objectively compare them (47, 48, 58, 59).
We applied an objective metric, the Matthew's Correlation Coefficient
(MCC), to evaluate numerous algorithms for clustering sequences into
OTUs. By performing this type of analysis, we were able to objectively
compare the algorithms, make recommendations to the field, and develop
new algorithms that outperformed the existing ones. Beyond evaluating
clustering algorithms, we have also evaluated methods of denoising
sequence data (60--62), assessed reference alignments (43, 44),
considered the importance of incorporating secondary structure
information in alignments (63), quantified the variation along the 16S
rRNA gene (44), and compared the statistical hypotheses tested by
commonly used tools (64). We have embraced the competition and diversity
of all methods being used to analyze amplicon data. This competition
forces us to identify the strengths and weaknesses of various methods so
that we can make recommendations to other researchers.

@@

leverage existing tools for other steps. Yet, we quickly appreciated the
need for providing other steps in a sequence analysis pipeline to make
other tools more accessible. This decision was motivated by learning
that the code for greengenes's (42) and ARB/SILVA's aligners were not
open source or publicly available. Thus, we realized that such an
important functionality needed to be opened to the community (43). More
recently, the rejection of closed source, commercial tools can be seen
by the broader adoption of open source alternatives. This has been the
case with the rising popularity of VSEARCH over USEARCH within the
microbial ecology community (55, 56). Related to insuring that mothur's
code is open source, our second core principle is that we maintain
transparency to our users. Perhaps a user does not need to interrogate
every line of code, but they need to understand what is happening. Many
programs, including online workflows, encapsulate large elements of a
pipeline in a single command. In contrast, mothur forces the user to
specify each step of the pipeline. Although the former approach makes an
analysis easier for a beginner, it stifles users that need greater
control or understanding of the assumptions at each step. This control
over the pipeline has made it easier for researchers to customize
databases or adapt the pipeline to analyze non-16S rRNA gene sequence
data. Furthermore, we have provided ample instructional materials to
teach users how to implement robust pipelines and the theory behind each
step. Third, as I mentioned above, there has been a plethora of methods
proposed for generating amplicon sequence data, and curating, aligning,
checking for chimeras, classifying, and clustering the data. I am proud
of the data-driven approach we have taken to comparing these methods. A
description of a new method has limited value if it is not benchmarked
against other methods or control datasets. Through this core principle
and mothur's large reach into the community, we have helped to develop
standards in the analysis of 16S rRNA gene sequence data. Fourth, a
focus on enabling reproducibility has always been central to the
functionality of mothur. From the beginning, mothur's logfiles have
represented a transcript of the user's command and outputs. When it
became clear that researchers were not submitting their sequence data to

@@

create a mothur command (make.sra) that creates a package for submitting

@@

generate. This has proven critical as it again represents transparency
and hopefully provides confidence to mothur's users that we are not
making recommendations that we do not follow ourselves.

\textbf{\emph{Challenges of making open source count.}} Anyone can post
code to GitHub with a permissive license and claim to be an open source
software developer. Far more challenging is engaging the target
community to make contributions to that code. Frankly, we have struggled
to expand the number of people that make contributions to the mothur
code base. One challenge we face is that if we looked to third parties
to contribute code to mothur, they would need to know C++. Given the
paucity of microbiologists that can program in a compiled language like
C++, expecting that community to provide contributors who can write code
in a syntax that prizes execution efficiency over developer efficiency
was not realistic. In contrast, the QIIME development team could be more
distributed because their code base was primarily written in Python,
which prizes developer efficiency over execution efficiency. QIIME is a
series of wrappers that allow users to execute other developers' code,
making the use of a scripting language like Python attractive. Their
choices resulted in many tradeoffs that have impacted ease of
installation, usability, execution speed, and flexibility. If we were
offered funding to rewrite mothur, we would likely rewrite it as an R
package that leaned heavily on the R language's C++ interface packages.
Of course, such choices are always best in hindsight. Yet, when we
started developing mothur, the ability to interface between scripting
languages like R and Python and C++ code was not as well developed as it
is today. For example, the modern version of the Rcpp package was first
released in 2009 and its popularity was not immediate (66). The
development of mothur has been a product of the environment that it was
created in. Although these decisions have largely had positive outcomes,
there have been tradeoffs that caused us to sacrifice other goals.

Beyond contributing to the mothur code base, we sought out other ways to
include the community as developers. The paper describing mothur
included 15 co-authors, most of whom responded to a call to provide a
wiki page that described how they used an early version of mothur to
analyze a data set. Our vision was that authors might use the mothur
wiki to document reproducible workflows for papers using mothur but to
also provide instructional materials for others seeking to adapt mothur
for their uses (\url{https://www.mothur.org/wiki}). Unfortunately, once
the incentive of co-authorship was removed, researchers stopped
contributing their workflows to the wiki. Again, this vision and the
lack of the community's adoption of wikis as a mechanism for reporting
workflows was a product of the environment. Although wikis were popular
in the late 2000's, they lacked the ability to directly execute the
commands that researchers reported. Such technology would not be
possible until the creation of IPython notebooks (2011) and R markdown
(2012). Another problem with the wiki approach was that potential
contributors did not see the wiki as a community resource. I frequently
received emails from scientists telling me that there was a typo on a
specific page when the intention was that they could correct the typos
without my input. We have been more successful in soliciting input and
contributions from the user community through the mothur discussion
forum and GitHub-based issue tracker. As mothur has matured, we have
been dependent on the user community to use these resources to tell us
what features they would like to see included in mothur and where the
documentation is confusing or incomplete
(\url{https://forum.mothur.org}). Often we can count on people not
directly affiliated with mothur to provide instruction and their own
experience to other users on the forum. We are constantly trying to
recruit our ``army'' and are happy to take any contributions we can.
Whether the contributions are to the code base, discussion forum, or
suggestions for new tools, these contributions have been invaluable to
the growth and popularity of mothur.

\textbf{\emph{Failed experiments.}} If we never failed, we would not be
trying hard enough. Over the past decade we have tried a number of
experiments to improve the usability and utility of mothur. One of our
first experiments was to use mothur to generate standard vector graphic
(SVG)-formatted files of heatmaps and Venn diagrams depicting the
overlap between microbial communities. Such visuals were helpful for
exploring data; however, I quickly realized that I would never put a
mothur-generated figure into a manuscript I wrote. Such visuals require
far too much customization to be publication-quality. Although QIIME has
incorporated visualization tools through the Emperor package (67), the
challenge of users taking default values has downsides, especially when
those defaults do not follow good data visualization principles. For
example, ordinations with black backgrounds and 3-D ordinations in a 2-D
medium now litter the literature. Instead, we have encouraged users to
use R packages to visualize mothur-generated results using the minimalR
instructional materials that I have developed
(\url{http://www.riffomonas.org/minimalR/}). A second experiment was the
creation of a graphical user interface (GUI) for running mothur. Forcing
users to interact with mothur through the command line has been a
significant hurdle for many. Unfortunately, the development effort
required to create and maintain a GUI is significant and there is
limited funding for such efforts. The newest version of QIIME (starting
with version 2.0.0) has emphasized interaction with the tools through a
GUI (68) and the related QIITA project offers a web-based GUI (69). It
remains to be seen how this experiment will go. Another downside of
using a GUI is that there is a risk that reproducibility will suffer if
users do not have a mechanism to document their mouse clicks.
Documentation of commands and parameter values is explicit in users can
provide a file with a list of commands and the software returns a
logfile with all commands and output recorded. Given the heightened
focus on reproducibility in recent years, we have extended significant
effort in developing instructional materials teaching users how to
organize, document, and execute reproducible pipelines that allow a user
to go from raw sequence data to a compiled manuscript (65, 70). A final
example of a failed experiment was a collaboration with programmers
through Google Summer of Code to develop commands in mothur that ran the
random forest and SVM machine learning algorithms. Similar to the
challenges of developing attractive visuals, fitting the algorithms'
hyperparameters, testing, and deploying the resulting models require a
significant amount of customization. Furthermore, machine learning is an
active area of research where methods are still being developed and
improved. Thankfully, there are numerous R and Python packages that do a
better job of developing these models (71, 72). In each of our
``failed'' experiments, the real problems were straying from what mothur
does well and failing to grasp what we really wanted the innovation to
do.

@@

The development of mothur would not be possible without the
contributions of its many supporters, developers, and users. Although
not a complete list, I would be remiss if I did not express my gratitude
to Sarah Westcott and the other members of my research group, Jo
Handelsman, Mitch Sogin, Susan Huse, Vincent Young, Lita Proctor, Kendra
Maas, and Marcy Balunas for their unique contributions to the continued
development of mothur.
